Faithful Model Explanations through\\ Energy-Constrained Conformal Counterfactuals\footnote{This is a pre-print.}

### 2023-12-19 experiments/notebooks/figure2.qmd
all charts in paper recreated

@@

    plts = []
    for target in 1:2
        X̂ = generate_conditional_samples(jem, _batch_size, target; niter=niter) 
        ex = extrema(hcat(X,X̂), dims=2)
        xlims = ex[1]
        ylims = ex[2]
        x1 = range(1.0f0.*xlims...,length=100)
        x2 = range(1.0f0.*ylims...,length=100)
        plt = Plots.contour(
            x1, x2, (x, y) -> softmax(jem([x, y]))[target], 
            fill=true, alpha=0.5, title="Target: $target", cbar=true,
            xlims=xlims,
            ylims=ylims,
        )
        Plots.scatter!(X[1,:], X[2,:], color=Int.(labels.refs), group=Int.(labels.refs), alpha=0.5)
        Plots.scatter!(
            X̂[1,:], X̂[2,:], 
            color=repeat([target], size(X̂,2)), 
            group=repeat([target], size(X̂,2)), 
            shape=:star5, ms=10
        )
        push!(plts, plt)
    end
    plt = Plots.plot(plts..., layout=(1, 2), size=(2*500, 400))
    display(plt)
end
```

@@

>>>>

This notebook can be used to recreate Figure 2 in the paper.

@@

    plts = []
    for target in 1:2
        X̂ = generate_conditional_samples(jem, _batch_size, target; niter=niter) 
        ex = extrema(hcat(X,X̂), dims=2)
        xlims = ex[1]
        ylims = ex[2]
        x1 = range(1.0f0.*xlims...,length=100)
        x2 = range(1.0f0.*ylims...,length=100)
        plt = Plots.contour(
            x1, x2, (x, y) -> softmax(jem([x, y]))[target], 
            fill=true, alpha=0.5, title="Target: $target", cbar=true,
            xlims=xlims,
            ylims=ylims,
        )
        Plots.scatter!(X[1,:], X[2,:], color=Int.(labels.refs), group=Int.(labels.refs), alpha=0.5)
        Plots.scatter!(
            X̂[1,:], X̂[2,:], 
            color=repeat([target], size(X̂,2)), 
            group=repeat([target], size(X̂,2)), 
            shape=:star5, ms=10
        )
        push!(plts, plt)
    end
    plt = Plots.plot(plts..., layout=(1, 2), size=(2*500, 400))
    display(plt)
end
savefig(plt, joinpath(output_images_path, "poc_conditional_samples.png"))
```

@@

The code below generates Figure 2 in the paper.

### 2023-12-19 paper/camera-ready/altmeyer.tex

Generating multiple samples using SGLD thus yields an empirical distribution $\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}$ that approximates what the model has learned about the input data. While in the context of EBM, this is usually done during training, we propose to repurpose this approach during inference in order to evaluate the faithfulness of model explanations. The appendix provides additional implementation details for any tasks related to energy-based modelling\footnote{The supplementary appendix can be found here: https://github.com/pat-alt/ECCCo.jl.}.

>>>>

Generating multiple samples using SGLD thus yields an empirical distribution $\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}$ that approximates what the model has learned about the input data. While in the context of EBM, this is usually done during training, we propose to repurpose this approach during inference in order to evaluate the faithfulness of model explanations. The appendix provides additional implementation details for any tasks related to energy-based modelling\footnote{The supplementary appendix can be found here: https://arxiv.org/abs/2312.10648.}.

### 2023-12-19 paper/contents/table_gen_params.tex

\caption{Generator hyperparameters: the optimiser step size ($\eta$); penalty strengths where $\lambda_1$ applies to all generators but \textit{Schut} and the other parameter are specific to \textit{ECCCo}; finally, the strength for the Ridge penalty on energy for \textit{ECCCo}.\label{tab:genparams} \newline}
\centering
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{rrrrrr}
\toprule
Dataset & $\eta$ & $\lambda_1$ & $\lambda_2$ & $\lambda_3$ & Ridge penalty\\
\midrule
Linearly Separable & 0.01 & 0.10 & 0.1 & 0.05 & 0.0\\
Moons & 0.01 & 0.10 & 0.1 & 0.50 & 0.0\\
Circles & 0.05 & 0.10 & 0.1 & 0.05 & 1.0\\
California Housing & 0.05 & 0.10 & 0.1 & 0.10 & 0.0\\
GMSC & 0.05 & 0.10 & 0.1 & 0.10 & 0.0\\
German Credit & 0.05 & 0.20 & 0.2 & 0.20 & 0.5\\
MNIST & 0.10 & 0.01 & 0.1 & 0.30 & 0.0\\
Fashion MNIST & 0.10 & 0.01 & 0.1 & 0.30 & 0.0\\
\bottomrule
\end{tabular}
\end{table}

>>>>

\caption{Generator hyperparameters: the optimiser step size ($\eta$); penalty strengths where $\lambda_1$ applies to all generators but \textit{Schut} and the other parameter are specific to \textit{ECCCo}; finally, the strength for the Ridge penalty on energy for \textit{ECCCo}.\label{tab:genparams} \newline}
\centering
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{rrrrrr}
\toprule
Dataset & $\eta$ & $\lambda_1$ & $\lambda_2$ & $\lambda_3$ & Ridge penalty\\
\midrule
Linearly Separable & 0.05 & 0.10 & 0.1 & 0.5 & 0.0\\
Moons & 0.05 & 0.10 & 0.1 & 0.2 & 0.0\\
Circles & 0.05 & 0.10 & 0.1 & 0.2 & 0.0\\
California Housing & 0.05 & 0.10 & 0.1 & 0.5 & 0.0\\
GMSC & 0.05 & 0.10 & 0.1 & 0.5 & 0.0\\
German Credit & 0.05 & 0.10 & 0.1 & 0.1 & 0.5\\
MNIST & 0.05 & 0.01 & 0.1 & 0.3 & 0.0\\
Fashion MNIST & 0.05 & 0.01 & 0.1 & 0.3 & 0.0\\
\bottomrule
\end{tabular}
\end{table}

### 2023-12-19 paper/partials/appendix.tex

\begin{figure}
  \centering
  \includegraphics[width=0.75\linewidth]{../www/mnist_generated_JEM Ensemble.png}
  \caption{Conditionally generated \textit{MNIST} images for our JEM Ensemble.}\label{fig:mnist-gen}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.5\linewidth]{../www/poc_conditional_samples.png}
  \caption{Conditionally generated samples (stars) for our \textit{Linearly Separable} data using a JEM.}\label{fig:poc-gen}
\end{figure}
\subsubsection{Inference: Quantifying Models' Generative Property}

@@

\begin{figure}
  \centering
  \includegraphics[width=1.0\linewidth]{../www/poc_set_size.png}
  \caption{Prediction set size (left), smooth set size loss (centre) and configurable classification loss (right) for a JEM trained on our \textit{Linearly Separable} data.}\label{fig:cp-diff}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=1.0\linewidth]{../www/mnist_all_lenet_eccco.png}
  \caption{Counterfactuals for \textit{MNIST} data generated by \textit{ECCCo+}. The underlying model is a LeNet-5 \textit{CNN}. Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-eccco-lenet}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=1.0\linewidth]{../www/mnist_all_mlp_eccco.png}
  \caption{Counterfactuals for \textit{MNIST} data generated by \textit{ECCCo+}. The underlying model is an \textit{MLP}. Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-eccco-mlp}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=1.0\linewidth]{../www/mnist_all_mlp_ens_eccco.png}
  \caption{Counterfactuals for \textit{MNIST} data generated by \textit{ECCCo+}. The underlying model is an \textit{MLP} ensemble. Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-eccco-mlp-ens}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=1.0\linewidth]{../www/mnist_all_jem_eccco.png}
  \caption{Counterfactuals for \textit{MNIST} data generated by \textit{ECCCo+}. The underlying model is a \textit{JEM}. Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-eccco-jem}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=1.0\linewidth]{../www/mnist_all_jem_ens_eccco.png}
  \caption{Counterfactuals for \textit{MNIST} data generated by \textit{ECCCo+}. The underlying model is a \textit{JEM} ensemble. Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-eccco-jem-ens}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=1.0\linewidth]{../www/mnist_all_lenet_revise.png}
  \caption{Counterfactuals for \textit{MNIST} data generated by \textit{REVISE}. The underlying model is a LeNet-5 \textit{CNN}. Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-revise-lenet}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=1.0\linewidth]{../www/mnist_all_mlp_revise.png}
  \caption{Counterfactuals for \textit{MNIST} data generated by \textit{REVISE}. The underlying model is an \textit{MLP}. Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-revise-mlp}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=1.0\linewidth]{../www/mnist_all_mlp_ens_revise.png}
  \caption{Counterfactuals for \textit{MNIST} data generated by \textit{REVISE}. The underlying model is an \textit{MLP} ensemble. Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-revise-mlp-ens}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=1.0\linewidth]{../www/mnist_all_jem_revise.png}
  \caption{Counterfactuals for \textit{MNIST} data generated by \textit{REVISE}. The underlying model is a \textit{JEM}. Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-revise-jem}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=1.0\linewidth]{../www/mnist_all_jem_ens_revise.png}
  \caption{Counterfactuals for \textit{MNIST} data generated by \textit{REVISE}. The underlying model is a \textit{JEM} ensemble. Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-revise-jem-ens}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=1.0\linewidth]{../www/vae_rec.png}
  \caption{Randomly drawn \textit{MNIST} images and their reconstructions generated by the VAE used by \textit{REVISE}.}\label{fig:vae-rec}
\end{figure}

>>>>

\begin{figure}
  \centering
  \includegraphics[width=0.75\linewidth]{www/mnist_generated_JEM_Ensemble.png}
  \caption{Conditionally generated \textit{MNIST} images for our JEM Ensemble.}\label{fig:mnist-gen}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.5\linewidth]{www/poc_conditional_samples.png}
  \caption{Conditionally generated samples (stars) for our \textit{Linearly Separable} data using a JEM.}\label{fig:poc-gen}
\end{figure}
\subsubsection{Inference: Quantifying Models' Generative Property}

@@

\begin{figure}
  \centering
  \includegraphics[width=1.0\linewidth]{www/poc_set_size.png}
  \caption{Prediction set size (left), smooth set size loss (centre) and configurable classification loss (right) for a JEM trained on our \textit{Linearly Separable} data.}\label{fig:cp-diff}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=1.0\linewidth]{www/mnist_all_lenet_eccco.png}
  \caption{Counterfactuals for \textit{MNIST} data generated by \textit{ECCCo+}. The underlying model is a LeNet-5 \textit{CNN}. Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-eccco-lenet}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=1.0\linewidth]{www/mnist_all_mlp_eccco.png}
  \caption{Counterfactuals for \textit{MNIST} data generated by \textit{ECCCo+}. The underlying model is an \textit{MLP}. Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-eccco-mlp}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=1.0\linewidth]{www/mnist_all_mlp_ens_eccco.png}
  \caption{Counterfactuals for \textit{MNIST} data generated by \textit{ECCCo+}. The underlying model is an \textit{MLP} ensemble. Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-eccco-mlp-ens}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=1.0\linewidth]{www/mnist_all_jem_eccco.png}
  \caption{Counterfactuals for \textit{MNIST} data generated by \textit{ECCCo+}. The underlying model is a \textit{JEM}. Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-eccco-jem}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=1.0\linewidth]{www/mnist_all_jem_ens_eccco.png}
  \caption{Counterfactuals for \textit{MNIST} data generated by \textit{ECCCo+}. The underlying model is a \textit{JEM} ensemble. Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-eccco-jem-ens}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=1.0\linewidth]{www/mnist_all_lenet_revise.png}
  \caption{Counterfactuals for \textit{MNIST} data generated by \textit{REVISE}. The underlying model is a LeNet-5 \textit{CNN}. Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-revise-lenet}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=1.0\linewidth]{www/mnist_all_mlp_revise.png}
  \caption{Counterfactuals for \textit{MNIST} data generated by \textit{REVISE}. The underlying model is an \textit{MLP}. Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-revise-mlp}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=1.0\linewidth]{www/mnist_all_mlp_ens_revise.png}
  \caption{Counterfactuals for \textit{MNIST} data generated by \textit{REVISE}. The underlying model is an \textit{MLP} ensemble. Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-revise-mlp-ens}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=1.0\linewidth]{www/mnist_all_jem_revise.png}
  \caption{Counterfactuals for \textit{MNIST} data generated by \textit{REVISE}. The underlying model is a \textit{JEM}. Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-revise-jem}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=1.0\linewidth]{www/mnist_all_jem_ens_revise.png}
  \caption{Counterfactuals for \textit{MNIST} data generated by \textit{REVISE}. The underlying model is a \textit{JEM} ensemble. Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-revise-jem-ens}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=1.0\linewidth]{www/vae_rec.png}
  \caption{Randomly drawn \textit{MNIST} images and their reconstructions generated by the VAE used by \textit{REVISE}.}\label{fig:vae-rec}
\end{figure}
